Cooperative Kernels: Safe Multitasking for Blocking Algorithms on GPUs

### 2016-11-15 write-ups/PLDI/paper.tex
section 5.5 getting shape

\begin{abstract}
There is growing interest in accelerating irregular data-parallel
algorithms on GPUs.  These algorithms are typically \emph{blocking},
so require fair scheduling.  But GPU programming models (e.g.\ OpenCL)
do not mandate fair scheduling, and GPU schedulers are inherently
unfair in practice.  Current approaches avoid this issue by exploiting
scheduling quirks particular to today's GPUs in a manner that does not
allow the GPU to be shared with other workloads (such as graphics
rendering tasks).  We propose \emph{cooperative kernels}, an extension
to the traditional GPU programming model geared towards writing
blocking algorithms.  Workgroups of a cooperative kernel \emph{are}
fairly scheduled, and multitasking is supported via a small set of
language extensions that the kernel and scheduler use to cooperate.
We describe the semantics of the cooperative kernels programming model
and our prototype implementation.  We evaluate the approach by porting
a set of irregular work stealing and graph algorithms to our
programming model and examining their performance under multitasking
across three GPUs.  Our implementation exploits no
vendor-specific hardware, driver or compiler support, so our encouraging
results provide a lower-bound on the efficiency with
which cooperative kernels can be implemented in practice.

@@

\paragraph{The needs of irregular data-parallel algorithms}
%Acceleration of general-purpose computations on graphics processing
%units (GPUs) has tended to focus on \emph{regular} data-parallel
%algorithms, for which work to be processed can be evenly split between
%the workgroups that execute a GPU kernel ahead of time.  However,
Many
interesting data-parallel algorithms are \emph{irregular}: the amount
of work to be processed is unknown ahead of time and may be determined
by the computation itself.  There is growing interest in accelerating such algorithms on GPUs, particularly algorithms that
traverse and manipulate linked data structures~\cite{owens-persistent,DBLP:conf/ipps/KaleemVPHP16,DBLP:conf/ipps/DavidsonBGO14,DBLP:conf/hipc/HarishN07,DBLP:journals/topc/MerrillGG15,DBLP:conf/egh/VineetHPN09,DBLP:conf/ppopp/NobariCKB12,DBLP:conf/hpcc/SolomonTT10a,DBLP:conf/popl/PrabhuRMH11,DBLP:conf/ppopp/Mendez-LojoBP12,DBLP:conf/oopsla/PaiP16,DBLP:conf/oopsla/SorensenDBGR16,dlb-web,TPO10,BNP12,Pannotia}.

@@

To operate correctly, a blocking concurrent algorithm requires
\emph{fair} scheduling of workgroups.  Without fairness, a
workgroup may fail to make progress, leading to starvation.  For
example, if one workgroup holds a mutex, an unfair scheduler may cause
another workgroup to spin-wait forever for the mutex to be
released.  Similarly, an unfair scheduler can cause a workgroup to spin-wait
indefinitely at a global barrier so that other workgroups do not reach the barrier.

\paragraph{A degree of fairness: occupancy-bound execution}
GPU schedulers are \emph{not} fair. The current GPU
programming models---OpenCL~\cite{opencl2Spec}, CUDA~\cite{cuda-75}  and
HSA~\cite{HSAprogramming11}---specify almost no guarantees regarding scheduling of
workgroups, and current GPUs do not
provide fair scheduling in practice.  Roughly speaking, each workgroup
executing a GPU kernel is mapped to a hardware \emph{compute
  unit}.\footnote{In practice, depending on the kernel, multiple workgroups might map to the same compute unit; we ignore this in our current discussion.}
%
The simplest way for a GPU driver to handle 
the case where more workgroups are launched than there are compute
units is via an \emph{occupancy-bound}
execution model~\cite{owens-persistent,DBLP:conf/oopsla/SorensenDBGR16}.  With this model, once a workgroup has commenced execution on a
compute unit (it has become \emph{occupant}), the workgroup has
exclusive access to the compute unit until it has finished execution.
Experiments suggest that the occupancy-bound execution model is widely
implemented by today's GPU devices and drivers~\cite{owens-persistent,DBLP:conf/oopsla/SorensenDBGR16,DBLP:conf/oopsla/PaiP16,BNP12}.   It is
evidently a simple model for device vendors to implement efficiently.

The occupancy-bound execution model does not guarantee fair scheduling
between workgroups: if all compute units are occupied then a
not-yet-occupant workgroup will not be scheduled until some occupant
workgroup completes execution.  Yet the execution model \emph{does}
provide fair scheduling between \emph{occupant} workgroups, which are
bound to separate compute units that operate in parallel.  Current GPU
implementations of blocking algorithms assume the
occupancy-bound execution model, which they exploit by launching 
no more workgroups than there are available compute
units~\cite{owens-persistent}. This works because \emph{today}'s GPUs seem to provide the
occupancy-bound execution model.

@@

Despite its practical prevalence, none of the current GPU programming
models actually mandate occupancy-bound execution.  Further, there are
clear reasons why this model is undesirable.
% from the point-of-view of
%GPU vendors and end users.
First, the execution model does not enable
multitasking, because a workgroup effectively \emph{owns} a compute
unit until the workgroup has completed execution.  The GPU cannot be used meanwhile for other
tasks, such as graphics rendering.
%Precisely because
%of this problem, operating systems employ a GPU \emph{watchdog} to
%heuristically detect and kill long-running computations on the GPU
%being used for display rendering~\cite{DBLP:conf/iwocl/SorensenD16}.
Second, \emph{energy throttling} is
an important concern for battery-powered devices~\cite{DBLP:journals/comsur/Vallina-RodriguezC13}.  In the future, it will be desirable for a mobile GPU driver to be able to
power down some compute units if the battery level is low or if
the user puts the device into a low power state.  Implementing energy
throttling would require being able to suspend the execution of a
workgroup, power down the compute unit to which the workgroup was
bound, and schedule the workgroup for later execution on a different
compute unit.

Our assessment, informed by discussions with a number of industrial
practitioners who have been involved in the OpenCL and/or HSA
standardisation efforts
(including~\cite{PersonalCommunicationRichards,PersonalCommunicationHowes}
and various engineers from GPU vendors), is that (1) for the above reasons, GPU vendors will
not commit to the occupancy-bound execution model that they currently
implement,
yet (2) GPU vendors will not guarantee fair scheduling, due to the
runtime overhead associated with supporting full preemption of
workgroups.  Vendors instead wish to
retain the essence of the simple occupancy-bound model, supporting preemption
only in key special cases.

\paragraph{Our proposal: cooperative kernels}
%
To summarise: blocking algorithms
demand fair scheduling of workgroups, but for good reasons
GPU vendors will not commit even to the guarantees of the
occupancy-bound execution model.

We propose \emph{cooperative kernels}, an extension to the GPU
programming model that aims to resolve this impasse.  A kernel
that requires fair scheduling is identified as \emph{cooperative}, and written using two additional
language primitives, $\offerkill$ and $\offerfork$, placed by the programmer.

At a point where the cooperative kernel could proceed with fewer
workgroups, a workgroup can execute $\offerkill$, offering to
sacrifice itself to the scheduler.  This indicates that the workgroup
would ideally continue executing, but that if required the
scheduler may destroy the workgroup for good.  The cooperative kernel
must be prepared to deal with either scenario.
%
At a point where the cooperative kernel could benefit from having
additional workgroups join the computation, a workgroup can execute
$\offerfork$.  This indicates that the cooperative
kernel is prepared to proceed with the existing set of workgroups, but
is able to take advantage of having one or more additional workgroups
commence execution directly after the $\offerfork$ program point.

@@

\vspace{-1mm}
\begin{enumerate}

@@

\item Cooperative kernels can be implemented over the occupancy-bound
  execution model that current GPUs provide: our proof-of-concept implementation requires
  no additional hardware or driver support.  Thus the model should be eay for vendors to implement directly.

@@

\end{enumerate}
\vspace{-1mm}

@@

\vspace{-1mm}
\begin{itemize}

@@

\end{itemize}
\vspace{-1mm}

We begin by providing background on OpenCL via two motivating examples (\mysec\ref{sec:background}).  At the end we discuss related work (\mysec\ref{sec:relatedwork}) and avenues for future work (\mysec\ref{sec:conclusion}).

@@

OpenCL specifically makes no guarantees about fair scheduling between
workgroups executing the same kernel, stating~\cite[p.\ 31]{opencl2Spec}: \emph{``A
  conforming implementation may choose to serialize the workgroups 
%so
%  a correct algorithm cannot assume that workgroups will execute in
%  parallel.  
\dots There is no safe and portable way to synchronize across
  the independent execution of workgroups since once in the work-pool,
  they can execute in any order.''}  CUDA similarly provides no guarantees~\cite{cuda-75}.

@@

\paragraph{Graph traversal example} \myfiglong\ref{fig:graphsearch} shows the
skeleton of a frontier-based graph traversal algorithm; such algorithms have
been shown to be execute efficiently on GPUs~\cite{BNP12,DBLP:conf/oopsla/PaiP16}. 
The kernel is
given three arguments in global memory: a graph structure, and two
arrays that can contain graph nodes. Initially, $\keyword{n0}$ contains the
starting nodes to process. Threads keep three local variables: $\keyword{level}$, which indicates the frontier level currently being
processed, and $\keyword{in\_nodes}$ and $\keyword{out\_nodes}$, which point to
distinct node arrays. The nodes to be
processed for the current frontier are contained in $\keyword{in\_nodes}$, while $\keyword{out\_nodes}$ contains
the nodes to be processed during the next frontier.

@@

\mytablong\ref{tab:applications} gives an overview of the 8 irregular
algorithms that we ported to cooperative kernels. Among them, 6 are
graph algorithms, based on Pannotia~\cite{Pannotia} and Lonestar~\cite{BNP12}
GPU application suites, using global barriers.  We indicate how many of
the original number of barriers are changed to resizing barriers, and
how many variables need to be transmitted.  In all cases all barriers
were converted, except for bc which contains barriers deeper in the call
stack, a case not supported by our prototype. Still, these barriers
could in principle also be converted. The remaining 2 algorithms are
workstealing applications: each required the addition of $\offerfork$
and $\offerkill$ at the start of the main loop, and no variables needed
to be transmitted.  For all these examples the porting to cooperative
kernels was similarly straightforward to the porting illustrated in
\mysec\ref{sec:programmingguidelines} for our running examples, and did
not require reshaping of the original algorithm. Most graph come with 2
different data sets, leading to 13 application/data set pairs in total.

@@

\ADComment{Preamble}

\HEComment{Below, just random notes so far}

Cooperative scheduling enables to concurrently run non-persistent
kernels alongside blocking ones. In terms of performance, we are
interested to see whether the variations of workgroups for the blocking
kernel.

Performance of cooperative scheduling corresponds to execution time of
cooperative kernels compared to the serial

In order to simulate non-persistent workloads alongside kernels of
blocking algorithms, we use a matrix multiplication kernel where
matrix dimensions are tuned to trigger the relevant execution
duration.

we compare :

time to execute original kernel

time to execute non-persistent tasks

time to execute both in cooperation

also, response time for non-persistent

use matmult to represent system workload, ref to previous section

\paragraph{Experimental setup}  \ADComment{Describe the data we are gathering.}

\paragraph{Results}  \ADComment{Discuss the results.}

@@

There has been a lot of work on accelerating blocking irregular
algorithms using GPUs, and on the \emph{persistent threads}
programming style for long-running kernels~\cite{owens-persistent,DBLP:conf/ipps/KaleemVPHP16,DBLP:conf/ipps/DavidsonBGO14,DBLP:conf/hipc/HarishN07,DBLP:journals/topc/MerrillGG15,DBLP:conf/egh/VineetHPN09,DBLP:conf/ppopp/NobariCKB12,DBLP:conf/hpcc/SolomonTT10a,DBLP:conf/popl/PrabhuRMH11,DBLP:conf/ppopp/Mendez-LojoBP12,DBLP:conf/oopsla/PaiP16,DBLP:conf/oopsla/SorensenDBGR16,dlb-web,TPO10,BNP12,Pannotia}.  These
approaches rely on the occupancy-bound execution model, and aim to
take ownership of the GPU by scheduling as many workgroups as the occupancy bound allows, relying on fair scheduling between these
workgroups.  This means the GPU is unavailable for other tasks, and
the method is not future-proof because (as discussed throughout the
paper) the occupancy-bound execution model is not guaranteed by GPU
programming models and is unlikely to be guaranteed in practice by
future GPU platforms.

@@

Hardware support for preemptive multitasking for \nvidia GPUs has been
proposed, enabling full preemption as well as \emph{SM-draining},
whereby workgroups occupying a symmetric multiprocessor (SM; a compute
unit using our terminology) are drained until the SM becomes free for
other tasks~\cite{DBLP:conf/isca/TanasicGCRNV14}.  A follow-up work adds the notion of SM \emph{flushing},
whereby a workgroup running on an SM can be cancelled and re-scheduled
from scratch if the workgroup has not yet committed any visible
side-effects to shared memory~\cite{DBLP:conf/asplos/ParkPM15}.  Both approaches have been evaluated
using simulators, over sets of regular GPU kernels.  State-of-the-art \nvidia GPUs targeting data centers (where multi-tasking is essential) do support
preemption, though it is not clear whether they guarantee fair scheduling~\cite{PascalWhitepaper}.

@@

Our solution does not impose a performance penalty on non-cooperative
kernels, which run in an unmodified manner, and careful placement of
our new language primitives can exploit the structure of a
cooperative kernel, enabling more efficient multitasking than
kernel-oblivious preemption.

@@

\paragraph{Dynamic parallelism} Recent versions of CUDA and OpenCL allow a kernel to spawn further kernels~\cite{cuda-75,opencl2Spec}.
This \emph{dynamic parallelism} can be exploited to implement a
GPU-based scheduler, by having an initial scheduler kernel repeatedly
spawn further kernels as required, according to some scheduling
policy~\cite{DBLP:conf/ppopp/Muyan-OzcelikO16}.  However, a kernel
that uses dynamic parallelism is still subject to the scheduling
limitations of current GPU programming models, and in particular the
problem that workgroups are not fairly scheduled.  Hence this feature
does not offer a solution to the problem of deploying blocking
algorithms on GPUs in a safe manner that we tackle with this work.

@@

Cooperative multitasking was offered in older operating systems
(e.g. Windows before the 95 edition) and is still used by some
operating systems, such as RISC
OS.\footnote{\url{http://www.riscos.info/index.php/Preemptive_multitasking}}
Many programming languages offer cooperative multitasking via
variations of coroutines, e.g. coroutine objects in
Python\footnote{\url{https://docs.python.org/3/reference/datamodel.html\#coroutine-objects}}
or Lua,\footnote{\url{https://www.lua.org/pil/9.1.html}} or Ruby's
\emph{fiber}.\footnote{\url{http://ruby-doc.org/core-2.1.1/Fiber.html}}

In the context of GPUs, cooperative multitasking has been used to
refer to the standard setup today, whereby the onus is on a GPU kernel
to complete execution within a reasonable time budget and return
control to the CPU application~\cite{adriaens2012case,CPE:CPE1722}.  In this setting, the programmer is responsible for ensuring cooperation between the host application and GPU kernel so that a kernel does not run for too long.
%
This is in contrast to our notion of cooperative kernels, which is designed specifically to support the needs of long-running kernels that one might wish to multitask with other workloads.

@@

\ADComment{Do citation tidy-up.}

@@

GPUs are known to have relaxed memory models~\cite{ABDGKPSW-2015}.  To abstractly
account for this, we assume that the shared state component $\sigma$
is not simplyl a mapping from locations to values, but instead
captures all the intricacies of GPU memory that can lead to this
relaxed behaviour.  We also assume a function $\sync$ which, given a
kernel state $\kappa$, returns a set of kernel states.  The idea is
that each $\kappa' \in \sync(\kappa)$ is a possible kernel state that
can be reached from $\kappa$ by stalling until all stores to memory
and loads from memory to thread-local storage have completed.  All we
require is that $\sync$ does not modify the number of active
workgroups nor the component of a thread state that determines which
statements remain to be executed.

>>>>

\begin{abstract}
There is growing interest in accelerating irregular data-parallel
algorithms on GPUs.  These algorithms are typically \emph{blocking},
so require fair scheduling.  But GPU programming models (e.g.\ OpenCL)
do not mandate fair scheduling, and GPU schedulers are inherently
unfair in practice.  This issue is currently circumvented by
exploiting scheduling quirks particular to today's GPUs in a manner
that does not allow the GPU to be shared with other workloads (such as
graphics rendering tasks).  We propose \emph{cooperative kernels}, an
extension to the traditional GPU programming model geared towards
writing blocking algorithms.  Workgroups of a cooperative kernel are
fairly scheduled, supporting blocking algorithms, and co-scheduling of
other tasks is supported via a small set of language extensions that
create a contract between the programmer and the scheduler.  We
describe the semantics of the cooperative kernels programming model
and discuss our portable prototype implementation.  We evaluate the
technique by porting a set of irregular work stealing and graph
algorithms to our programming model and assessing the performance and
responsiveness of the approach across several GPUs.  Because our
implementation exploits no vendor-specific hardware, driver or
compiler support, our encouraging experimental results provide a
lower-bound on the efficiency with which cooperative kernels can be
implemented in practice.

@@

\paragraph{The needs of irregular data-parallel algorithms}
%Acceleration of general-purpose computations on graphics processing
%units (GPUs) has tended to focus on \emph{regular} data-parallel
%algorithms, for which work to be processed can be evenly split between
%the workgroups that execute a GPU kernel ahead of time.  However,
Many
interesting data-parallel algorithms are \emph{irregular}: the amount
of work to be processed is unknown ahead of time and may be determined
by the computation itself.  The last decade has seen growing interest in accelerating irregular
data-parallel algorithms on GPUs, particularly algorithms that
traverse and manipulate linked data structures~\cite{owens-persistent,TPO10,DBLP:conf/ipps/KaleemVPHP16,DBLP:conf/ipps/DavidsonBGO14,DBLP:conf/hipc/HarishN07,DBLP:journals/topc/MerrillGG15,DBLP:conf/egh/VineetHPN09,DBLP:conf/ppopp/NobariCKB12,DBLP:conf/hpcc/SolomonTT10a,DBLP:conf/popl/PrabhuRMH11,DBLP:conf/ppopp/Mendez-LojoBP12,DBLP:conf/oopsla/PaiP16,DBLP:conf/ipps/KaleemVPHP16,DBLP:conf/oopsla/SorensenDBGR16,dlb-web,TPO10}.

@@

To operate correctly, a blocking concurrent algorithm requires
\emph{fair} scheduling, which in the context of GPUs means fair scheduling of workgroups.  Without fairness, a
workgroup may fail to make progress, leading to starvation.  For
example, if one workgroup holds a mutex, an unfair scheduler may cause
another workgroup to spin-wait forever for the mutex to be
released.  Similarly, an unfair scheduler can cause a workgroup to spin-wait
indefinitely at a global barrier so that other workgroups do not reach the barrier.

\paragraph{A degree of fairness: occupancy-bound execution}
GPU schedulers are \emph{not} fair. the current GPU
programming models---OpenCL~\cite{opencl2Spec}, \ADComment{Tyler, please add best CUDA since I've asked you to add it below in any case} CUDA~\cite{...}  and
HSA~\cite{HSAprogramming11}---specify almost no guarantees regarding scheduling of
workgroups, and current GPUs do not
provide fair scheduling in practice.  Roughly speaking, each workgroup
executing a GPU kernel is mapped to a hardware \emph{compute
  unit}.\footnote{In practice the situation can be more complex: if a
  kernel is sufficiently simple and if workgroups are small enough,
  multiple workgroups may be mapped to a single compute unit.  We assume, for the purpose of the current discussion, that the kernel is sufficiently complex that this does not happen.}
%
The simplest way for a GPU driver to handle 
the case where more workgroups are launched than there are compute
units is via an \emph{occupancy-bound}
execution model~\cite{DBLP:conf/oopsla/SorensenDBGR16}. \ADComment{Tyler, please add best cite for OB model; I guess our OOPSLA paper introduced the term, but we should cite that + Owens?}  With this model, once a workgroup has commenced execution on a
compute unit (it has become \emph{occupant}), the workgroup has
exclusive access to the compute unit until it has finished execution,
after which a subsequent workgroup can be bound to the compute unit.
Experiments suggest that the occupancy-bound execution model is widely
implemented by today's GPU devices and drivers~\cite{DBLP:conf/oopsla/SorensenDBGR16}. \ADComment{For the previous cite, should we cite other papers as well as ours?}  It is
evidently a simple model for device vendors to implement efficiently.

The occupancy-bound execution model does not guarantee fair scheduling
between workgroups: if all compute units are occupied then a
not-yet-occupant workgroup will not be scheduled until some occupant
workgroup completes execution.  Yet the execution model \emph{does}
provide fair scheduling between \emph{occupant} workgroups, which are
bound to separate compute units that operate in parallel.  Current GPU
implementations of irregular data parallel algorithms assume the
occupancy-bound execution model, which they exploit by launching 
no more workgroups than there are available compute
units~\cite{...}. \ADComment{Tyler, please pick best cite for previous
  claim.} This works because \emph{today}'s GPUs seem to provide the
occupancy-bound execution model.

@@

Despite its practical prevalence, none of the current GPU programming
models actually mandate occupancy-bound execution.  Further, there are
clear reasons why this model is undesirable.
% from the point-of-view of
%GPU vendors and end users.
First, the execution model does not enable
multitasking, because a workgroup effectively \emph{owns} a compute
unit until the workgroup has completed execution.  If a long-running
kernel occupies all compute units, the GPU cannot be used for other
tasks, such as graphics rendering, in the meantime.
%Precisely because
%of this problem, operating systems employ a GPU \emph{watchdog} to
%heuristically detect and kill long-running computations on the GPU
%being used for display rendering~\cite{DBLP:conf/iwocl/SorensenD16}.
Second, \emph{energy throttling} is
an important concern for battery-powered devices, such as smartphones
and tablets~\cite{DBLP:journals/comsur/Vallina-RodriguezC13}.  In the future, it will be desirable for a mobile GPU driver to be able to
power down one or more compute units if the battery level is low or if
the user puts the device into a low power state.  Implementing energy
throttling would require being able to suspend the execution of a
workgroup, power down the compute unit to which the workgroup was
bound, and schedule the workgroup for later execution on a different
compute unit.

Our assessment, informed by discussions with a number of industrial
practitioners who have been involved in the OpenCL and/or HSA
standardisation efforts
(including~\cite{PersonalCommunicationRichards,PersonalCommunicationHowes}
and various engineers from GPU vendors), is that (1) GPU vendors will
not commit to the occupancy-bound execution model that they currently
implement, due to the need for multitasking and energy throttling,
yet (2) GPU vendors will not guarantee fair scheduling, due to the
runtime overhead associated with supporting full preemption of
workgroups~\cite{ISCAPAPERSreeMentioned}.  Vendors instead wish to
retain the essence of the occupancy-bound model, which is
simple to implement efficiently, supporting preemption
only in key special cases.

\paragraph{Our proposal: cooperative kernels}
%
To summarise: blocking algorithms, such as most irregular algorithms,
demand a degree of fair scheduling of workgroups, but for good reasons
GPU vendors will not commit to the guarantees of even the
occupancy-bound execution model.

We propose \emph{cooperative kernels}, an extension to the GPU
programming model that aims to resolve this impasse.  A GPU kernel
that requires workgroups to be fairly scheduled is identified as a
\emph{cooperative} kernel, and is written using two additional
language primitives, $\offerkill$ and $\offerfork$, placed by the programmer.

At a point where the cooperative kernel could proceed with fewer
workgroups, a workgroup can execute $\offerkill$, offering to
sacrifice itself to the scheduler.  This indicates that the workgroup
would ideally continue with its execution but that, if required, the
scheduler may destroy the workgroup for good: the cooperative kernel
is prepared to deal with either scenario.

At a point where the cooperative kernel could benefit from having
additional workgroups join the computation, a workgroup can execute
$\offerfork$.  This indicates that the cooperative
kernel is prepared to proceed with the existing set of workgroups, but
is able to take advantage of having one or more additional workgroups
commence execution directly after the $\offerfork$ program point.

@@

\begin{enumerate}

@@

\item Cooperative kernels can be implemented over the occupancy-bound
  execution model that current GPUs provide---we present a
  proof-of-concept implementation that requires
  no special-purpose hardware or driver support.

@@

\end{enumerate}

@@

\begin{itemize}

@@

\end{itemize}

We begin by providing an overview of the OpenCL programming model and
two motivating examples of irregular algorithms (\mysec\ref{sec:background}).  At the end of the paper we discuss related work (\mysec\ref{sec:relatedwork}) and avenues for future work (\mysec\ref{sec:conclusion}).

@@

OpenCL specifically makes no guarantees about fair scheduling between
workgroups executing the same kernel, stating~\cite[p.\ 31]{opencl2Spec}: \emph{``A
  conforming implementation may choose to serialize the workgroups 
%so
%  a correct algorithm cannot assume that workgroups will execute in
%  parallel.  
\dots There is no safe and portable way to synchronize across
  the independent execution of workgroups since once in the work-pool,
  they can execute in any order.''}  \ADComment{Tyler: please provide CUDA reference with page} CUDA similarly provides no guarantees~\cite{...}.

@@

\paragraph{Graph traversal example} \myfiglong\ref{fig:graphsearch} shows the
skeleton of a frontier-based graph traversal algorithm; such algorithms have
been shown to be execute efficiently on GPUs~\cite{...}. \ADComment{Tyler, can you please work out which of the graph algorithm papers is the right one to cite here?}
The kernel is
given three arguments in global memory: a graph structure, and two
arrays that can contain graph nodes. Initially, $\keyword{n0}$ contains the
starting nodes to process. Threads keep three local variables: $\keyword{level}$, which indicates the frontier level currently being
processed, and $\keyword{in\_nodes}$ and $\keyword{out\_nodes}$, which point to
distinct node arrays. The nodes to be
processed for the current frontier are contained in $\keyword{in\_nodes}$, while $\keyword{out\_nodes}$ contains
the nodes to be processed during the next frontier.

@@

\mytablong\ref{tab:applications} gives an overview of the 8 irregular
algorithms that we ported to cooperative kernels. Among them, 6 are
graph algorithms, based on Pannotia~\cite{...} and Lonestar~\cite{...}
GPU application suites, using global barriers.  We indicate how many of
the original number of barriers are changed to resizing barriers, and
how many variables need to be transmitted.  In all cases all barriers
were converted, except for bc which contains barriers deeper in the call
stack, a case not supported by our prototype. Still, these barriers
could in principle also be converted. The remaining 2 algorithms are
workstealing applications: each required the addition of $\offerfork$
and $\offerkill$ at the start of the main loop, and no variables needed
to be transmitted.  For all these examples the porting to cooperative
kernels was similarly straightforward to the porting illustrated in
\mysec\ref{sec:programmingguidelines} for our running examples, and did
not require reshaping of the original algorithm. Most graph come with 2
different data sets, leading to 13 application/data set pairs in total.

@@

Cooperative scheduling enables to run a non-cooperative kernel alongside
a cooperative one, we focus here on the performance impact of this
multitasking and on the scheduling responsiveness. The performance is
measured with respect to the performance of each kernel run in
sequence. Responsiveness, i.e., how fast a non-cooperative kernel can be
scheduled, depends not only on the scheduler implementation, but also on
how often the cooperative kernel offer workgroups to be killed.

% In order to simulate non-persistent workloads alongside kernels of
% blocking algorithms, we use a matrix multiplication kernel where
% matrix dimensions are tuned to trigger the relevant execution
% duration.

\paragraph{Experimental setup}
For a given cooperative kernel and its data set, we make it run while
regularly asking to schedule a non-cooperative kernel that simulates
graphics rendering, according to the classes listed in
\mytab~\ref{tab:noncooperativeworkload}. In practice, we use a matrix
multiplication kernel as the non-cooperative workload, with matrix
dimensions tailored for each GPU to reach the appropriate execution
duration.  In separate runs, we ask the scheduler to launch this
non-cooperative kernels with different number of workgroups: either 1, a
quarter, a half or all but one workgroups among the ones available for
the cooperative kernel. Moreover, for graph algorithms we try both the
regular and the query barrier.

Our experiments span 13 pairs of cooperative kernels and data sets, 3
classes of non-cooperative kernel workloads and frequency, 4 quantities
of workgroups claimed for the non-cooperative kernel and 2 variations of
resizing barriers for graph algorithms, leading to 288
configurations. Each of this configuration is run 10 times on each
GPU. For each run, we record the execution time of the cooperative
kernel. For each scheduling of the non-cooperative kernel during the
run, we also record the \emph{gather time} needed by the scheduler to
collect enough workgroups to launch the non-cooperative kernel, and the
non-cooperative kernel execution time.

Whenever a cooperative or non-cooperative kernel terminates, we verify
that it has really performed its computation by comparing its results
against the expected one. We encountered two kind of issues during the
experiments: some configurations would lead to a system freeze or crash
although they run fine on other GPUs. Although we don't claim that our
programs are all correct, this issue seems to indicate a bug in the
OpenCL compiler or driver of the problematic GPU. We also have some
cooperative kernels that returned invalid results when the
non-cooperative kernel takes all but one of the cooperative kernel
workgroups, we decided to ignore these results.

\paragraph{Results}  \ADComment{Discuss the results.}
explain horizontal lines

missing results for all but one in bfs rmat22 since errors

also, high gather time corresponds to applications and data sets where
resize/kill is triggered less often.

Each configuration was run 10 times on each GPU.

@@

There has been a lot of work on accelerating blocking irregular
algorithms using GPUs~\cite{...}, and on the \emph{persistent threads}
programming style for long-running kernels~\cite{...}.  These
approaches rely on the occupancy-bound execution model, and aim to
take ownership of the GPU by scheduling as many workgroups as the GPU
has compute units, relying on fair scheduling between these
workgroups.  This means the GPU is unavailable for other tasks, and
the method is not future-proof because (as discussed throughout the
paper) the occupancy-bound execution model is not guaranteed by GPU
programming models and is unlikely to be guaranteed in practice by
future GPU platforms.

@@

Hardware support for preemptive multitasking for \nvidia GPUs has been
proposed, enabling full preemption as well as \emph{SM-draining},
whereby workgroups occupying a symmetric multiprocessor (SM; a compute
unit using our terminology) are drained until the SM becomes free for
other tasks.  A follow-up work adds the notion of SM \emph{flushing},
whereby a workgroup running on an SM can be cancelled and re-scheduled
from scratch if the workgroup has not yet committed any visible
side-effects to shared memory.  Both approaches have been evaluated
using simulators, over sets of regular GPU kernels.

@@

Our solution does not impose a performance penalty on non-cooperative
kernels, which run in an unmodified manner, and careful placement of
our new language primitives that exploit the structure of a
cooperative kernel may enable more efficient multitasking than
kernel-oblivious preemption.

@@

\paragraph{Dynamic parallelism} Recent version of CUDA and OpenCL allow a kernel to spawn further kernels~\cite{...}.
This \emph{dynamic parallelism} can be exploited to implement a GPU-based scheduler: an initial scheduler kernel uses the feature to spawn additional kernels as required~\cite{ }.

 is launched

%
\ADComment{Kepler preemption support?  Mention in related work?}

\ADComment{Sreepathi mentioned the NCCL framework.}

\cite{TPO10}

Dynamic parallelism could be extended to give barriers, but doesn't work with work stealing (less general).  It is in recent OpenCL.

\cite{} \ADComment{Basic idea: map tasks to
  SMs, not to workgroups, so that tasks that would benefit from being
  processed by the same SM can be scheduled to the same SM.  Has
  little to do with preemption, but more to do with irregular
  computation.}

\cite{DBLP:conf/isca/TanasicGCRNV14} \ADComment{I have not read, but
  it's about enabling preemptive multiprogramming, so we should read
  it carefully.}

\cite{DBLP:conf/ppopp/Muyan-OzcelikO16} \ADComment{I have not read; we
  should understand this one carefully.}

@@

\HEComment{cooperative multitasking: OS: RISC OS, IBM CICS, Windows
  before 95, old MacOS, maybe Oberon? TRON?. In GPU context: some
  paper mention ``cooperative'' multitasking but they simply refer to
  the fact that kernels should not compute for too long in order to
  let other have access to it. Also, many prog languages implements
  variations of coroutines, We can see blocking channels of CSP (or
  Golang more recently) as yielding (sending a message is yielding
  computation to the process which receives it)}

Cooperative multitasking was offered in ancient operating systems
(e.g. Windows before the 95 edition) and is still used nowadays in
RISC OS for instance. Many programming languages offer cooperative
multitasking via variations of coroutines, e.g. coroutine objects in
Python\footnote{\url{https://docs.python.org/3/reference/datamodel.html\#coroutine-objects}}
or Lua\footnote{\url{https://www.lua.org/pil/9.1.html}}, or Ruby's
fiber \footnote{\url{http://ruby-doc.org/core-2.1.1/Fiber.html}}. In
the context of GPU resources, cooperative multitasking seems to refer
to sizing kernels such that they do not compute for too long in order
to yield the whole GPU regularly~\cite{adriaens2012case}. Our
contribution present a much more flexible cooperative multitasking,
where the long-running application is never totally stopped.

\HEComment{ RISC OS see :
\url{http://www.riscos.info/index.php/Preemptive_multitasking}
}

% ###
% Cooperative multitasking for GPU-accelerated grid systems

% http://onlinelibrary.wiley.com/doi/10.1002/cpe.1722/full

% @article {CPE:CPE1722,
% author = {Ino, Fumihiko and Ogita, Akihiro and Oita, Kentaro and Hagihara, Kenichi},
% title = {Cooperative multitasking for GPU-accelerated grid systems},
% journal = {Concurrency and Computation: Practice and Experience},
% volume = {24},
% number = {1},
% publisher = {John Wiley & Sons, Ltd},
% issn = {1532-0634},
% url = {http://dx.doi.org/10.1002/cpe.1722},
% doi = {10.1002/cpe.1722},
% pages = {96--107},
% keywords = {multitasking, GPU, CUDA, grid computing},
% year = {2012},
% }

% Not super interesting: they just divide blocking computation in
% smaller independent kernels, and fires them at regular intervals to
% let the system be responsive.

@@

@@

GPUs are known to have relaxed memory models~\cite{ }.  To abstractly
account for this, we assume that the shared state component $\sigma$
is not simplyl a mapping from locations to values, but instead
captures all the intricacies of GPU memory that can lead to this
relaxed behaviour.  We also assume a function $\sync$ which, given a
kernel state $\kappa$, returns a set of kernel states.  The idea is
that each $\kappa' \in \sync(\kappa)$ is a possible kernel state that
can be reached from $\kappa$ by stalling until all stores to memory
and loads from memory to thread-local storage have completed.  All we
require is that $\sync$ does not modify the number of active
workgroups nor the component of a thread state that determines which
statements remain to be executed.
